Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-02 doc/paper/paper.tex
Update more sections of paper
Updates from Overleaf

\usepackage{arxiv}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{lipsum}
\usepackage{caption}
\usepackage{float}

@@

% TODO: Lyapunov calculation

@@

@@

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|}
         % TODO: \lambda_1?
         % TODO: calculated values
        \hline
        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda$ & calculated $\lambda$ & relative error of calculated $\lambda$ \\
        \hline \hline
        16 & 45.92 & 4 & 1.50255 & & \\
        16 & 40 & 4 & 1.37446 & & \\
        10 & 28 & 8/3 & 0.90566 & & \\
        \hline
    \end{tabular}
    \caption{
        Lorenz Parameters and Largest Lyapunov Exponent
        \cite{viswanath1998lyapunov}
    }
    \label{table:lorenz_params}
\end{table}

@@

@@

@@

@@

Repeating the same experiment, but with different parameters, the results are similar (see 
Figure \ref{fig:doub_pend_energy}).

\begin{figure}[H]
    \centering
    %\includegraphics[.5\linewidth]{images/chaos_vs_energy_in_doub_pend.png}
    \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}
    \label{fig:doub_pend_energy}
\end{figure}

>>>>

\usepackage{arxiv}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{amsmath}
\usepackage{lipsum}
\usepackage{caption}
\usepackage{float}
\usepackage{graphicx}

% the comment 'DATA POINT' will mark where data was calculated and then inserted into
% the paper.

@@

% TODO: Lyapunov calculation
% TODO: Test with Circle yielding zero

@@

% TODO: numerical solutions section

@@

% TODO: Image of Lorenz System

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|}
         % TODO: \lambda_1?
         % TODO: calculated values
        \hline
        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda$ & calculated $\lambda$ & relative error of calculated $\lambda$ \\
        \hline \hline
        16 & 45.92 & 4 & 1.50255 & & \\ % DATA POINTS
        16 & 40 & 4 & 1.37446 & & \\
        10 & 28 & 8/3 & 0.90566 & & \\
        \hline
    \end{tabular}
    \caption{
        Lorenz Parameters and Largest Lyapunov Exponent
        \cite{viswanath1998lyapunov}
    }
    \label{table:lorenz_params}
\end{table}

@@

The 3-body problem, or the generalized n-body problem, is a famous problem which many have 
analyzed; most notably, Herni Poincaré wrote many papers developing the area significantly
\cite{chenciner2000remarkable}. One of Poincaré's solutions, known as the circular 
restricted 3-body problem, demonstrates chaotic properties \cite{oestreicher2007history}. 
Poincaré performs several reductions of the original 3-body problem in order to come to a 
much more useful form of the problem. The mass of one of the objects is assumed to be 
negligible. This is the case for Sun-Earth-Moon orbits. The moon has little effect on the 
orbit of the sun. Another aspect is the rotating reference frame. Since the mass of one of the bodies is negligible, the problem simplifies to first solving a two body problem. While much easier to solver, Poincaré went further to stop the motion of the two planets by using a rotating reference frame. Rotating with the planets, the problem can be modeled seemingly by two static planets and a third mass-less planet orbiting around those in a two dimensional plane.

% norm from https://tex.stackexchange.com/questions/107186/how-to-write-norm-which-adjusts-its-size
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\rv}{\mathbf{r}}
The N body problem is defined as follows \cite{chenciner2000remarkable}:

Let $\rv_i$ be the position of the $i^\text{th}$ body in the N body problem.
Then, the force of gravity is the sum of gravity from every other body in the system.
That is,

\begin{equation}
   F_g^{(i)} = m_i \frac{\mathrm{d}^2 \rv_i}{\mathrm{dt}^2} =
   \sum_{j \neq i} G \frac{m_i m_j \hat{\rv}_{ij}}{\norm{\rv_{ij}}^3}
   \label{eq:nbody}
\end{equation}

where $\rv_{ij} = \rv_j - \rv_i$, $m_i$ is the mass of body $i$, and $G$ is the gravitational constant.

For our purposes, the reduced system is more useful since it is much easier to calculate than numerically solving the original 3-body problem. With three bodies in three dimension, their motion is described by 9 second order differential equations. Therefore, reducing the system to first order differential equations yields 18 first order equations. The reduced problem only requires 4 first order differential equations since the position is defined by two second order differential equations.

Derived in \cite{eberle2007case}, the system of equations defined with dimensionless rotating (synodic) coordinates:

\begin{align}
    x'' - 2 y' &= x
        - \frac{\alpha}{r_1^3} (x - \mu)
        - \frac{\mu}{r_2^3} (x + \alpha) \nonumber \\
    y'' + 2 x' &= \left(
        1 - \frac{\alpha}{r_1^3}
        - \frac{\mu}{r_2^3}
    \right) y \nonumber \\
\end{align}

where

\begin{align}
    \mu &= \frac{m_1}{m_1 + m_2} \nonumber \\
    \alpha &= 1 - \mu \nonumber \\
    r_1(t) &= \left[(x(t) - \mu)^2 + y(t)^2\right]^{\frac{1}{2}} \nonumber \\
    r_2(t) &= \left[(x(t) + \alpha)^2 + y(t)^2\right]^{\frac{1}{2}} \label{eq:reduc_3body_cor}
\end{align}

and where mass body 1 and two are located at $-\alpha$ and $\mu$.

If we define the following system:

\begin{align}
    x_1 &= x \nonumber \\
    x_2 &= x' \nonumber \\
    y_1 &= y \nonumber \\
    y_2 &= y', \label{eq:first_order_def}
\end{align}

then, we can take the derivative of each:

\begin{align*}
    x_1' &= x' \\
    &= x_2,
\end{align*}
and
\begin{align*}
    x_2' &= x'' \\
    &= 2 y' + x
        - \frac{\alpha}{r_1^3} (x - \mu)
        - \frac{\mu}{r_2^3} (x + \alpha) \\
    &= 2 y_2 + x_1 - \frac{\alpha}{r_1^3} (x_1 - \mu)
        - \frac{\mu}{r_2^3} (x_1 + \alpha),
\end{align*}
and
\begin{align*}
    y_1' &= y' \\
    &= y_2,
\end{align*}
and
\begin{align*}
    y_2' &= y'' \\
    &= - 2 x' + \left(
        1 - \frac{\alpha}{r_1^3}
        - \frac{\mu}{r_2^3}
    \right) y \\
    &= - 2 x_2 + \left(
        1 - \frac{\alpha}{r_1^3}
        - \frac{\mu}{r_2^3}
    \right) y_1.
\end{align*}

Then, we can represent the second order system of motion as four first order equations:

\begin{align}
    x_1' &= x_2 \nonumber \\
    x_2' &= 2 y_2 + x_1 - \frac{\alpha}{r_1^3} (x_1 - \mu)
        - \frac{\mu}{r_2^3} (x_1 + \alpha) \nonumber \\
    y_1' &= y_2 \nonumber \\
    y_2' &= - 2 x_2 + \left(
        1 - \frac{\alpha}{r_1^3}
        - \frac{\mu}{r_2^3}
    \right) y_1, \label{eq:reduce_3_body_prog_sys} \\
\end{align}

where

\begin{align}
    r_1 &= \left[(x_1 - \mu)^2 + y_1^2\right]^{\frac{1}{2}} \nonumber \\
    r_2 &= \left[(x_1 + \alpha)^2 + y_1^2\right]^{\frac{1}{2}}.
\end{align}

Here, $x_1$ and $y_1$ represent the position of the third body relative to the other two,
but not in the standard 
Cartesian coordinates of 3D space. Outlined in the paper \cite{eberle2007case}, the 
transformation back to Cartesian coordinates is given, but we don't need the exact 
position of the bodies since we are only concerned with the chaotic motion, so that isn't 
shown here.

There are however two issues when integrating this problem. As objects approach each
other, the distance between them 
approaches zero. If we have a collision, then this approaches infinity. There is no 
collision detection in this model as we assume falsely that each body has a radius of zero,
so this case is not handled. Instead, we will get a 
divide by zero error, and the model cannot be calculated further. Further, near collisions 
are also an issue. Detailed in \cite{chambers1999hybrid}, with the distance between bodies 
approaching zero, their potential energy becomes significantly smaller than other values, 
such as those describing kinetic energy. This 
produces a stiff system of equations, and introduces error. Since energy is conserved in
this system (as it is a Hamiltonian system), the error introduced represents energy being 
created. This causes the singularity property seen in Figure \ref{fig:singularity}

\begin{figure}[h]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision.png}
        \caption{The restricted 3 body problem with collision}
        \label{fig:r3b_collision}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{images/r3b_collision_whole.png}
        \caption{The resultant divergence of the body}
        \label{fig:r3b_collision_whole}
    \end{subfigure}
    \caption{The Double Pendulum}
    \label{fig:doub_pend}
\end{figure}

While it is possible to solve this issue with symplectic integrators \cite{chambers1999hybrid},

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=.5\linewidth]{images/example_doub_pend.png}
    \caption{The Double Pendulum}
    \label{fig:doub_pend}
\end{figure}

@@

$\theta_1$ and $\omega_1$ represent the angle and angular velocity of the inner pendulum 
where $\theta_1=0$ implies the pendulum is straight down, and $\theta_1=\epsilon$ implies 
the pendulum is rotated $\epsilon$ radians counter-clockwise. The same is true for the 
outer pendulum relative to its pivot point, the end of the first pendulum. $l_1$ and $l_2$ 
are the lengths of the inner and outer pendulums, respectively. $m_1$ and $m_2$ are the 
masses of the balls on the end points of the inner and outer pendulums, respectively.
Then, the positions of the endpoints $(x_1$, $y_1)$ and $(x_2, y_2)$ of the pendulum, with 
the center pivot at the origin, $(0, 0)$, can be calculated as follows:

\begin{align}
    x_1 &= l_1 \sin(\theta_1), \nonumber \\
    y_1 &= - l_1 \cos(\theta_1), \nonumber \\
    x_2 &= x_1 + l_2 \sin(\theta_2), \nonumber \\
    y_2 &= y_1 - l_2 \cos(\theta_2), \label{eq:doub_pend_endpoints}
\end{align}

% TODO: position of the pendulum end points

@@

Repeating the same experiment, but with different parameters ($l_1 = l_2 = 1$, and
$m_1 = m_2 = 1$), the results are similar (see Figure \ref{fig:doub_pend_energy}).
As inner theta increases, the largest LCE of the system seems to
approach 1.4. % DATA POINT

\begin{figure}[H]
    \centering
    \includegraphics[width=.5\linewidth]{images/chaos_vs_energy_in_doub_pend.png}
    \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}
    \label{fig:doub_pend_energy}
\end{figure}
